Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-10-08 prop_letter.tex
prop_letter.tex \section{Born} Updating...

Recall in Sec.~\ref{sec:fuzzy}, we illustrate the non-constructive
postulates of the quantum probability measures first, and then introduce
the Born rule which induces a quantum probability measures from each
state. When we shift our attention to QIVPMs, we also want to construct
QIVPMs according to a simple rule. Since the ranges of quantum probability
measures and QIVPMs are $\left[0,1\right]$ and $\mathscr{I}$, respectively,
if there is a function~$\iota$ mapping a number~$x\in\left[0,1\right]$
to an interval~$\iota\left(x\right)\in\mathscr{I}$, then we can
concatenate it with a quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
and get a $\mathscr{I}$-valued function~$\bar{\mu}:\events\overset{\mu}{\rightarrow}\left[0,1\right]\overset{\iota}{\rightarrow}\mathscr{I}$.
Although $\bar{\mu}$ and QIVPMs have the same domain and range, $\bar{\mu}$
is not necessary a QIVPM. For example, the following theorem highlight
a situation where $\bar{\mu}$ cannot be a QIVPM.

\begin{thm}Given a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$,
if there exists three numbers~$x_{0}$, $x_{1}$, and $x_{2}\in\left[0,1\right]$
such that $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$, but 
\begin{equation}
\iota\left(x_{0}+x_{1}+x_{2}\right)+\iota\left(x_{2}\right)\nsubseteq\iota\left(x_{0}+x_{2}\right)+\iota\left(x_{1}+x_{2}\right)\,.\label{eq:iota-Inclusion-not}
\end{equation}
Then, there is a quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
such that $\iota\circ\mu$ is not a QIVPM, where $\circ$ is the function
concatenation.\end{thm}

Consider the state
\begin{equation}
\begin{aligned}\rho={} & x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}\\
 & +\left(1-x_{0}-x_{1}-x_{2}\right)\proj{3}\,.
\end{aligned}
\end{equation}
Then, the function~$\iota\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}:\events\rightarrow\mathscr{I}$
fails to satisfy Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
\begin{equation}
\begin{aligned} & \mathrel{\hphantom{\nsubseteq}}\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{0}+\proj{1}+\proj{2}\right)\right)+\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{2}\right)\right)\\
 & \nsubseteq\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{0}+\proj{2}\right)\right)+\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{1}+\proj{2}\right)\right)\,,
\end{aligned}
\end{equation}
and cannot be a QIVPM.

In other words, even if $\iota\circ\mu$ has the same domain and range
as a QIVPM, we may need to verify whether $\iota\circ\mu$ is a QIVPM
individually. For example, in the four-dimensional Hilbert space,
\begin{equation}
\iota\left(x\right)=\begin{cases}
\imposs\,, & \textrm{ if }x=0\,;\\
\left[0,\tfrac{1}{2}\right]\,, & \textrm{ if }x\in\left(0,\tfrac{1}{4}\right]\,;\\
\unknown\,, & \textrm{ if }x\in\left(\tfrac{1}{4},\tfrac{3}{4}\right)\,;\\
\left[\tfrac{1}{2},1\right]\,, & \textrm{ if }x\in\left[\tfrac{3}{4},1\right)\,;\\
\necess\,, & \textrm{ if }x=1\,.
\end{cases}
\end{equation}
and $\rho=\left(\proj{0}+\proj{1}+\proj{2}+\proj{3}\right)/4$ inducing
a QIVPM~$\iota\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}$ is a non-trivial
fact~\citep{HOSTunpublished}. In order to verify whether $\iota\circ\mu$
is a QIVPM easier, we can impose some conditions on $\iota$. Notice
that our counterexamples are in a four-dimensional Hilbert space,
the conditions for $\iota$ for a Hilbert space of dimension~$d\le3$
is not as strict as the general conditions.

\begin{thm}In a Hilbert space of dimension~$d\le3$, if $\iota:\left[0,1\right]\rightarrow\mathscr{I}$
subject to the following constraints:\begin{subequations}\label{eq:iota-constraints}
\begin{eqnarray}
 &  & \iota(0)=\imposs\,,\\
 &  & \iota(1)=\necess\,,\\
 &  & \iota\left(1-x\right)=\necess-\iota\left(x\right)\,,
\end{eqnarray}
\end{subequations}and satisfying for any three numbers~$x_{0}$
and $x_{1}\in\left[0,1\right]$, $x_{0}+x_{1}\in\left[0,1\right]$
implies
\begin{equation}
\iota_{n}\left(x_{0}+x_{1}\right)\subseteq\iota_{n}\left(x_{0}\right)+\iota_{n}\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
Then, $\iota\circ\mu:\events\rightarrow\mathscr{I}$ is always a QIVPM
for any quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$.\end{thm}

Although their concatenation $\iota\circ\mu$ is always a function
from $\events$ to $\mathscr{I}$, to guarantee $\iota\circ\mu$ is
a QIVPM, $\iota$ needs to satisfy some extra conditions.

\begin{thm}If $\iota\circ\mu:\events\rightarrow\mathscr{I}$ is always
a QIVPM for any quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
in any Hilbert space $\Hilb$, then $\iota:\left[0,1\right]\rightarrow\mathscr{I}$
must subject to the following constraints:\begin{subequations}
\begin{eqnarray}
 &  & \iota(0)=\imposs\,,\\
 &  & \iota(1)=\necess\,,\\
 &  & \iota\left(1-x\right)=\necess-\iota\left(x\right)\,,
\end{eqnarray}
\end{subequations} and satisfying for any three numbers~$x_{0}$,
$x_{1}$, and $x_{2}\in\left[0,1\right]$, $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$
implies
\begin{equation}
\iota\left(x_{0}+x_{1}+x_{2}\right)+\iota\left(x_{2}\right)\subseteq\iota\left(x_{0}+x_{2}\right)+\iota\left(x_{1}+x_{2}\right)\,.\label{eq:iota-Inclusion}
\end{equation}
\end{thm}

Although a interval-valued function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
may look exotic, its left-end and right-end~$\left[\iotal\left(x\right),\iotar\left(x\right)\right]=\iota\left(x\right)$
are just usual functions from a unit interval to another unit interval.
Eq.~(\ref{eq:iota-Inclusion}) then implies that $\iotal$ is Wright-convex,
i.e., $\iotal\left(x_{0}+x_{1}+x_{2}\right)+\iotal\left(x_{2}\right)\ge\iotal\left(x_{0}+x_{2}\right)+\iotal\left(x_{1}+x_{2}\right)$,
and $\iotar$ is Wright-concave, i.e., $\iotar\left(x_{0}+x_{1}+x_{2}\right)+\iotar\left(x_{2}\right)\le\iotar\left(x_{0}+x_{2}\right)+\iotar\left(x_{1}+x_{2}\right)$~\citep{Wright1954,RobertsVarberg1973,PecaricTong1992,Brzdek_2014}.
Notice that a Wright-convex function is midpoint-convex or Jensen-convex,
i.e., 
\begin{equation}
\frac{\iotal\left(x_{0}\right)+\iotal\left(x_{1}\right)}{2}\ge\iotal\left(\frac{x_{0}+x_{1}}{2}\right)\,,
\end{equation}
and a midpoint-convex function is continuous on an open interval if
we cannot use the Axiom of Choices~\citep{Blumberg_1919,Sierpinski1920,RobertsVarberg1973}.
Therefore, the range of $\iota$ either is $\left\{ \necess,\imposs,\unknown\right\} $
or contains infinitely many intervals.

@@

To prove that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$ is a QIVPM,
it is sufficient to check Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}).
Given $x_{0}$, $x_{1}$, and $x_{0}+x_{1}\in\left[0,1\right]$,
\begin{equation}
\iota_{n}\left(x_{0}+x_{1}\right)\subseteq\iota_{n}\left(x_{0}\right)+\iota_{n}\left(x_{1}\right)\label{eq:iota-add}
\end{equation}
implies Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}$ and $P_{1}$ are orthogonal. When $P_{0}$ and $P_{1}$
are not orthogonal on a three-dimensional Hilbert space, we have two
possibilities:
\begin{itemize}
\item Their complements~$\mathbb{1}-P_{0}$ and $\mathbb{1}-P_{1}$ are
orthogonal so that we can apply Eq.~(\ref{eq:iota-add}) on their
complements.
\item One of them projects onto a subspace of what the other one projects
onto.
\end{itemize}
\noindent Both cases deduce Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
easily~\citep{HOSTunpublished} so that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
is indeed a QIVPM.

>>>>

Recall in Sec.~\ref{sec:fuzzy}, we illustrate the non-constructive
postulates of the quantum probability measures first, and then introduce
the Born rule which induces a quantum probability measures from each
state. When we shift our attention to QIVPMs, we also want to construct
QIVPMs according to a simple rule. Since the ranges of quantum probability
measures and QIVPMs are $\left[0,1\right]$ and $\mathscr{I}$, respectively,
if there is a function~$\iota$ mapping a number~$x\in\left[0,1\right]$
to an interval~$\iota\left(x\right)\in\mathscr{I}$, then we can
concatenate it with a quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
and get a $\mathscr{I}$-valued function~$\bar{\mu}:\events\overset{\mu}{\rightarrow}\left[0,1\right]\overset{\iota}{\rightarrow}\mathscr{I}$.
Although $\bar{\mu}$ and QIVPMs have the same domain and range, $\bar{\mu}$
is not necessary a QIVPM. For example, in the four-dimensional Hilbert
space with 
\begin{equation}
\iota'\left(x\right)=\begin{cases}
\imposs\,, & \textrm{ if }x=0\,;\\
\left[0,\tfrac{1}{2}\right]\,, & \textrm{ if }x\in\left(0,\tfrac{1}{4}\right]\,;\\
\unknown\,, & \textrm{ if }x\in\left(\tfrac{1}{4},\tfrac{3}{4}\right)\,;\\
\left[\tfrac{1}{2},1\right]\,, & \textrm{ if }x\in\left[\tfrac{3}{4},1\right)\,;\\
\necess\,, & \textrm{ if }x=1\,,
\end{cases}\label{eq:iota}
\end{equation}
the function~$\iota'\circ\mu_{\rho_{0}}^{\mathrm{\mathrm{B}}}$ induced
by $\rho_{0}=\frac{1}{4}\cdot\proj{0}+\frac{1}{4}\cdot\proj{1}+\frac{1}{4}\cdot\proj{2}+\frac{1}{4}\cdot\proj{3}$
is a QIVPM, but the function~$\iota'\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$
induced by $\rho_{1}=\frac{1}{8}\cdot\proj{0}+\frac{1}{8}\cdot\proj{1}+\frac{1}{8}\cdot\proj{2}+\frac{5}{8}\cdot\proj{3}$
is \emph{not} a QIVPM, where $\circ$ is the function concatenation.
In particular, verifying whether an \emph{arbitrary} function is a
QIVPM is a long and tedious process~\citep{HOSTunpublished}, so
that it would be good if $\iota\circ\mu$ is automatically a QIVPM
during some situations.

\begin{thm}\label{thm:QRVPM2QIVPM-3}In a Hilbert space of dimension~$d\le3$,
if $\iota:\left[0,1\right]\rightarrow\mathscr{I}$ subject to the
following constraints:\begin{subequations}\label{eq:iota-constraints}
\begin{eqnarray}
 &  & \iota(0)=\imposs\,,\\
 &  & \iota(1)=\necess\,,\\
 &  & \iota\left(1-x\right)=\necess-\iota\left(x\right)\,,
\end{eqnarray}
\end{subequations}and satisfying for any three numbers~$x_{0}$
and $x_{1}\in\left[0,1\right]$, $x_{0}+x_{1}\in\left[0,1\right]$
implies
\begin{equation}
\iota\left(x_{0}+x_{1}\right)\subseteq\iota\left(x_{0}\right)+\iota\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
Then, $\iota\circ\mu:\events\rightarrow\mathscr{I}$ is always a QIVPM
for any quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$.\end{thm}

The situation for $d=2$ is trivial. When $d=3$, to prove that $\iota\circ\mu$
is a QIVPM, it is sufficient to check Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}).
When $P_{0}$ and $P_{1}$ are orthogonal, Eq.~(\ref{eq:iota-add})
implies Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}).
When $P_{0}$ and $P_{1}$ are not orthogonal on a three-dimensional
Hilbert space, we have two possibilities:
\begin{itemize}
\item Their complements~$\mathbb{1}-P_{0}$ and $\mathbb{1}-P_{1}$ are
orthogonal so that we can apply Eq.~(\ref{eq:iota-add}) on their
complements.
\item One of them projects onto a subspace of what the other one projects
onto.
\end{itemize}
\noindent Both cases deduce Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
easily~\citep{HOSTunpublished} so that $\iota\circ\mu$ is indeed
a QIVPM.

Since $\iota'$ defined in Eq.~(\ref{eq:iota}) satisfies Eq.~(\ref{eq:iota-constraints})
and Eq.~(\ref{eq:iota-add}), every function~$\iota'\circ\mu$ on
a three-dimensional Hilbert space is a QIVPM according to Thm.~\ref{thm:QRVPM2QIVPM-3}.
However, we knows that $\iota'\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$
is not a QIVPM. This highlight the necessity of the conditions to
guarantee $\iota\circ\mu$ is always a QIVPM.

\begin{thm}If $\iota\circ\mu:\events\rightarrow\mathscr{I}$ is always
a QIVPM for any quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
in any Hilbert space $\Hilb$, then $\iota:\left[0,1\right]\rightarrow\mathscr{I}$
must subject to Eqs.~(\ref{eq:iota-constraints}) and satisfying
for any three numbers~$x_{0}$, $x_{1}$, and $x_{2}\in\left[0,1\right]$,
$x_{0}+x_{1}+x_{2}\in\left[0,1\right]$ implies
\begin{equation}
\iota\left(x_{0}+x_{1}+x_{2}\right)+\iota\left(x_{2}\right)\subseteq\iota\left(x_{0}+x_{2}\right)+\iota\left(x_{1}+x_{2}\right)\,.\label{eq:iota-Inclusion}
\end{equation}
\end{thm}

Suppose $\iota$ doesn't satisfy Eq.~(\ref{eq:iota-Inclusion}),
there must exist three numbers~$x_{0}$, $x_{1}$, and $x_{2}\in\left[0,1\right]$
such that $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$, but 
\[
\iota\left(x_{0}+x_{1}+x_{2}\right)+\iota\left(x_{2}\right)\nsubseteq\iota\left(x_{0}+x_{2}\right)+\iota\left(x_{1}+x_{2}\right)\,.
\]
Consider the state
\begin{equation}
\begin{aligned}\rho={} & x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}\\
 & +\left(1-x_{0}-x_{1}-x_{2}\right)\proj{3}\,.
\end{aligned}
\end{equation}
Then, the function~$\iota\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}:\events\rightarrow\mathscr{I}$
fails to satisfy Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
\[
\begin{aligned} & \mathrel{\hphantom{\nsubseteq}}\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{0}+\proj{1}+\proj{2}\right)\right)+\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{2}\right)\right)\\
 & \nsubseteq\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{0}+\proj{2}\right)\right)+\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{1}+\proj{2}\right)\right)\,,
\end{aligned}
\]
and cannot be a QIVPM. Therefore, to make sure $\iota\circ\mu$ is
always a QIVPM, $\iota:\left[0,1\right]\rightarrow\mathscr{I}$ should
satisfy Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-Inclusion}).

Since $\iota'\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$ is not a
QIVPM, we know $\iota'$ doesn't satisfy Eq.~(\ref{eq:iota-Inclusion}).
In order to find a function~$\iota$ satisfies Eq.~(\ref{eq:iota-Inclusion}),
we break down into two functions: its left-end and right-end~$\left[\iotal\left(x\right),\iotar\left(x\right)\right]=\iota\left(x\right)$.
They are both usual functions from a unit interval to another unit
interval. Eq.~(\ref{eq:iota-Inclusion}) then implies that $\iotal$
is Wright-convex, i.e., $\iotal\left(x_{0}+x_{1}+x_{2}\right)+\iotal\left(x_{2}\right)\ge\iotal\left(x_{0}+x_{2}\right)+\iotal\left(x_{1}+x_{2}\right)$,
and $\iotar$ is Wright-concave, i.e., $\iotar\left(x_{0}+x_{1}+x_{2}\right)+\iotar\left(x_{2}\right)\le\iotar\left(x_{0}+x_{2}\right)+\iotar\left(x_{1}+x_{2}\right)$~\citep{Wright1954,RobertsVarberg1973,PecaricTong1992,Brzdek_2014}.
Notice that a Wright-convex function is midpoint-convex or Jensen-convex,
i.e., 
\begin{equation}
\frac{\iotal\left(x_{0}\right)+\iotal\left(x_{1}\right)}{2}\ge\iotal\left(\frac{x_{0}+x_{1}}{2}\right)\,,
\end{equation}
and a midpoint-convex function is continuous on an open interval if
we cannot use the Axiom of Choices~\citep{Blumberg_1919,Sierpinski1920,RobertsVarberg1973}.
Therefore, the range of $\iota$ either is $\left\{ \necess,\imposs,\unknown\right\} $
or contains infinitely many intervals.

@@

To prove that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$ is a QIVPM,
it is sufficient to check Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}).
Given $x_{0}$, $x_{1}$, and $x_{0}+x_{1}\in\left[0,1\right]$,
\begin{equation}
\iota_{n}\left(x_{0}+x_{1}\right)\subseteq\iota_{n}\left(x_{0}\right)+\iota_{n}\left(x_{1}\right)\label{eq:iota-add-1}
\end{equation}
implies Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}$ and $P_{1}$ are orthogonal. When $P_{0}$ and $P_{1}$
are not orthogonal on a three-dimensional Hilbert space, we have two
possibilities:
\begin{itemize}
\item Their complements~$\mathbb{1}-P_{0}$ and $\mathbb{1}-P_{1}$ are
orthogonal so that we can apply Eq.~(\ref{eq:iota-add-1}) on their
complements.
\item One of them projects onto a subspace of what the other one projects
onto.
\end{itemize}
\noindent Both cases deduce Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
easily~\citep{HOSTunpublished} so that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
is indeed a QIVPM.
